---
title: "Humanitarian Data"
format:   
html:
    code-fold: true
    code-tools: true
editor_options: 
  chunk_output_type: console
---

### Visualizing Humanitarian Data

This document includes code on visualizing the aid that landed in "backward" countries during the COVID-19 pandemic.

### Data:

Data comes from Kaggle.

### Libraries

```{r results='hide'}

library(tidyverse)
library(xda) # a good package for data exploration
library(hrbrthemes)
library(ggthemes)
library(ggplot2)
library(patchwork)
library(thematic)
```

```{r echo = TRUE}
country_data <- read.csv("C:\\Users\\HP\\Downloads\\archive\\Country-data.csv", stringsAsFactors = F)
```

Now, we explore the data using head(). Alternatively, tail() can be used. In addition, specifying a number \[e.g., head(vector, 10)\] will show top 10 observations.

```{r}
head(country_data)
```

It is a good practice to observe the type of data using str().

```{r}
str(country_data)
```

Descriptive statistics also give a good picture of what we're dealing with.

```{r}
numSummary(country_data)
```

### Exploratory visualizations

We use ggplot2 to create some visualizations to explore the structure of our data. For instance,

```{r}
country_data %>% 
  select(country,life_expec) %>% 
  arrange(life_expec) %>% 
  slice(1:10) %>% 
  ggplot(aes(x = reorder(country, -life_expec), y =  life_expec)) +
  geom_bar(stat = 'identity',
           width = 0.4) +
  theme_void() +
  geom_col(fill = c("#E70F15")) +
  geom_text(aes(label = life_expec), size = 3,
            fontface = 'bold', hjust = 1.5,
            color = 'white') +
  theme(
    axis.text.y = element_text(),
  plot.subtitle = element_text(size = 10)
  ) +
  labs(title = 'Which Countries have the shortest Life Expectancy',
       subtitle = 'Humanitarian intervention amid COVID-19 pandemic face unexpected challenges in these countries with already low life spans',
       x = element_blank(),
       y = 'Life Span in years',
       caption = 'Source = HELP International') +
  coord_flip() 
```

How much are bottom 20 countries by gdpp spending on health?

```{r echo=FALSE}
p1 <- country_data %>%
  select(gdpp, health) |>
  arrange(gdpp) |>
  slice(1:20) |>
  ggplot(aes(health)) +
  geom_boxplot()
  labs(x = '% of GDP',
       y = element_blank())

p2 <- country_data %>%
  select(gdpp, health) |>
  arrange(desc(gdpp)) |>
  slice(1:20) |>
  ggplot(aes(health)) +
  geom_boxplot() +
  labs(x = '% of GDP',
       y = element_blank())

# combining plots using patchwork
p1 + p2 + plot_annotation('How do countries with lowest and highest GDPP relate in Health Spending?', subtitle = 'Top 20 countries on the left; bottom 20 countries on the right',theme=theme(plot.title=element_text(hjust=0)))
```

If COVID-19 interventions were to aim to increase % health spending in lower spending countries, how will it impact life expectancy?

We compare 20 lowest % GDP spenders on health with the top 20 % GDP spenders on health.

```{r}
country_data %>% 
  select(country, health, life_expec) %>% 
  arrange(health) %>% 
  slice(1:20) %>% 
  ggplot(aes(health, life_expec)) +
  geom_point() +
  geom_smooth(method = lm) -> p3

country_data %>% 
  select(country, health, life_expec) %>% 
  arrange(desc(health)) %>% 
  slice(1:20) %>% 
  ggplot(aes(health, life_expec)) +
  geom_point() +
  geom_smooth(method = lm) -> p4

p3 + p4 + plot_annotation('% GDP spending on health is weakly correlated with increase in life expectancy', subtitle = 'Compared to 20 lowest % health spending countries (left), 20 highest spenders have touched the point of diminishing return')
```
